---
title: "High COVID19 Death Rates in Indiana, Michigan, and Ohio"
author: "Jacob Moore"
output: pdf_document
date: "2022-10-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

The objective of this study is to present and analyze COVID19 data for the US states of Indiana, Michigan, and Ohio, in order to determine if, despite their proximity, there may be some non-geographical variable which caused a notable difference in either case numbers or deaths. 

This study will first present case and death numbers for the US as a whole in order to contextualize the state-level data. Then, the individual data for each state is presented and compared.

All written code and console outputs are preserved within this report in order to facilitate their examination.

```{r libraries}
library(tidyverse)
library(lubridate)
```

```{r get_jhu_data}
### Get current Data in the four files
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c("time_series_covid19_confirmed_US.csv",
                "time_series_covid19_deaths_US.csv")

urls <- str_c(url_in, file_names)
```

```{r import_data, messsage = FALSE}

US_cases <- read_csv(urls[1], show_col_types = FALSE)
US_deaths <- read_csv((urls[2]), show_col_types = FALSE)

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <- read_csv(uid_lookup_url, show_col_types = FALSE) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))
```

```{r tidy_data}

US_cases <- US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

US_deaths <- US_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

US <- US_cases %>%
  full_join(US_deaths)

US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

US_by_state <- US_by_state %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

US_totals <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  ungroup()

US_totals <- US_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

```

## US Data

Below is the visualization of total COVID19 cases and deaths. Note that the visualization is presented on a logarithmic scale to preserve legibility. We can see that while growth continues to the present in both cases and deaths, growth overall has not reached the frighteningly rapid pace of 2020 in either 2021 or 2022.

Further below is the visualization for new cases and deaths in the US as a whole. We can see, as expected from the visualization of totals, that new cases level off in mid-2020, though not without some spikes, especially in the winter months.

```{r US_visualizations}

# graphs for US
US_totals %>%
  filter(cases > 0 ) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) + 
  geom_point(aes(y = deaths, color = "deaths")) + 
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "Total COVID19 Cases and Deaths in US", y=NULL)

US_totals %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) + 
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "New COVID19 Cases and Deaths in US", y=NULL)

```

## Best and Worst Death Rates in US States and Territories

Below is a table showing the states and territories within the US with the lowest death rate as measured by deaths per 1000 people in the population. Below that is a similar table showing those states with the highest death rate. Note the presence of Michigan in the second table as the state with the tenth highest death rate. Note also that death rates and case rates appear to correlate.

```{r US_analysis}
#transform 'by state' data
US_state_totals <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000 * cases / population,
            deaths_per_thou = 1000 * deaths / population) %>%
  filter(cases > 0, population > 0)

# 10 best states
US_state_totals %>%
  slice_min(deaths_per_thou, n = 10) %>%
  select(deaths_per_thou, cases_per_thou, everything())

# 10 worst states
US_state_totals %>%
  slice_max(deaths_per_thou, n = 10) %>%
  select(deaths_per_thou, cases_per_thou, everything())
```

## Modeling Death Rates and Case Rates

The model below formalizes the apparent relationship between death rate and case rate. We can see a clear correlation but the relationship is, of course, not perfectly linear.

```{r US_model}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)
summary(mod)
US_state_totals_w_pred <- US_state_totals %>% mutate(pred = predict(mod))
US_state_totals_w_pred %>% ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red") 
  
```

## Regional Clusters

When we isolate only those states in which we are interested, Indiana, Michigan, and Ohio, we can see that all three have more deaths than predicted by our model. We also see that Ohio has fewer cases, but performs no better with respect to deaths.

These results may point to some shared regional factor or factors that increase the risk of death for those who contracted the disease.

```{r my_states_model}

# add model that only includes my states!
my_state_totals_w_pred <- US_state_totals_w_pred %>% filter(Province_State == "Michigan" | Province_State == "Indiana" | Province_State == "Ohio" )

my_state_totals_w_pred %>% ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red") +
  geom_text(aes(x = cases_per_thou, y = deaths_per_thou, label = Province_State), nudge_x = -.5, nudge_y = .05)

```

Other regional clusters of states do not have this same tendency. For example, in New England, if we look at New York, Connecticut, and Rhode Island:

```{r new_england_states_model}

# add model that only includes my states!
new_england_state_totals_w_pred <- US_state_totals_w_pred %>% filter(Province_State == "New York" | Province_State == "New Hampshire" | Province_State == "Connecticut" )

new_england_state_totals_w_pred %>% ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red") +
  geom_text(aes(x = cases_per_thou, y = deaths_per_thou, label = Province_State), nudge_x = .7, nudge_y = .05)

```

If we look to the Pacific Northwest in Washington, Oregon, and Idaho:

```{r northwest_states_model}

# add model that only includes my states!
northwest_state_totals_w_pred <- US_state_totals_w_pred %>% filter(Province_State == "Washington" | Province_State == "Oregon" | Province_State == "Idaho" )

northwest_state_totals_w_pred %>% ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red") +
  geom_text(aes(x = cases_per_thou, y = deaths_per_thou, label = Province_State), nudge_x = -.5, nudge_y = .05)

```

Finally, looking at the Southwestern states of New Mexico, Utah, and Arizona:

```{r southwest_states_model}

# add model that only includes my states!
southwest_state_totals_w_pred <- US_state_totals_w_pred %>% filter(Province_State == "New Mexico" | Province_State == "Arizona" | Province_State == "Utah" )

southwest_state_totals_w_pred %>% ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red") +
  geom_text(aes(x = cases_per_thou, y = deaths_per_thou, label = Province_State), nudge_x = .75, nudge_y = .1)

```

Clearly, none of these other regions show the same sort of tendency toward a clustering of death rates. The Midwestern tri-state area of Ohio, Indiana, and Michigan is unique in the way that they all exceed predicted death rates.

## State Level Data for Indiana

Looking at the states individually we see that, despite the high death rates, each follows the relative national trends in deaths and cases relatively closely. First, we have total and new deaths and cases for Indiana.

```{r Indiana_visualizations}

US_by_state %>%
  filter(Province_State == "Indiana") %>%
  filter(cases > 0 ) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) + 
  geom_point(aes(y = deaths, color = "deaths")) + 
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "Total COVID19 Cases and Deaths in Indiana", y=NULL)

US_by_state %>%
  filter(Province_State == "Michigan") %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) + 
  geom_point(aes(y = new_deaths, color = "new_deaths")) + 
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "New COVID19 Cases and Deaths in Indiana", y=NULL)

```

## State Level Data for Michigan

The same visualizations for Michigan show much the same pattern.

```{r Michigan_visualizations}
# graphs for Michigan
US_by_state %>%
  filter(Province_State == "Michigan") %>%
  filter(cases > 0 ) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) + 
  geom_point(aes(y = deaths, color = "deaths")) + 
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "Total COVID19 Cases and Deaths in Michigan", y=NULL)

US_by_state %>%
  filter(Province_State == "Michigan") %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) + 
  geom_point(aes(y = new_deaths, color = "new_deaths")) + 
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "New COVID19 Cases and Deaths in Michigan", y=NULL)
```

## State Level Data for Ohio

Finally, Ohio follows our pattern as well.

```{r Ohio_visualizations}

US_by_state %>%
  filter(Province_State == "Ohio") %>%
  filter(cases > 0 ) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) + 
  geom_point(aes(y = deaths, color = "deaths")) + 
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "Total COVID19 Cases and Deaths in Ohio", y=NULL)

US_by_state %>%
  filter(Province_State == "Ohio") %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) + 
  geom_point(aes(y = new_deaths, color = "new_deaths")) + 
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "New COVID19 Cases and Deaths in Ohio", y=NULL)

```

## Biases

The collected data has several possible sources of bias. We can see that the states vary dramatically in frequency of dates with 0 new cases. This may be the result of a lack of regular reporting. This irregularity may be a confounding factor in other respects as well.

The collected data also only reflects reported cases and deaths. The federal government no longer mandates data reporting in the same way as early in the pandemic. This may lead us to believe that more recent data is less reliable.

## Conclusion

We can see that while Indiana, Michigan, and Ohio follow, in general, the same general trends in case numbers and deaths as the rest of the United States, the death rate is higher than we would expect based on our model. This may point to issues in regional healthcare infrastructure or regional cultural make up that leads to distinctive lifestyle factors.

## Suggestions for Further Research

Additional research ought to be conducted to analyse the relationship in this region between COVID deaths and other factors to determine which, if any, are playing a decisive role in these outcomes.

This is imperative to prevent possible deaths as we continue to deal with the ongoing COVID situation and for any possible future pandemics or epidemics that may effect the region.
